---
title: "Deliverables"
author: "EJ Arce"
date: "5/4/2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library("RCurl")
library("dplyr")
library("ggplot2")
library("tidyr")
```


## Forecasting California legislative election results

  Forecasting elections is critical to the campaign strategies of candidates running for office. For example, a US presidential candidate needs 270 electoral votes to win an election, so they must efficiently allocate campaign resources to ensure victories in as many states as possible. Thus, it is important to know in which states the race is tight so they can focus campaign efforts on those states, while ignoring states that do not include tight races. Considering that much of the forecasting field has focused on US elections, our project focuses on forecasting election results for the California state legislature.

  There are two types of state legislative elections: state senatorial elections and house elections.  The state of California (along with every other state) is divided up into two types of districts for these elections: assembly and senatorial districts. California has 80 assembly (house) districts and 40 senatorial districts. Each assembly district elects an assemblymember every two years and each senatorial district elects a senator every four years. Like some state national elections (e.g. US senatorial and gubernatorial elections), these elections take place every November of an even year.  However, unlike some national elections (i.e. the presidential campaigns), the popular vote of the given district determines the winner of the race and because the political campaign takes place within a single district, there is not as big a need to allocate resources per se or focus on areas where the votes/delegates could go either way. However, these elections are not so simple...
  
  The boundaries of both the state and assembly districts are redrawn every ten years according to Article XXI in the California Constitution. While this may not play a major role in districts with fairly consistent voting tendencies, this boundary changing may be highly influential in areas with random dispersals of Republicans and Democrats (or Purple districts). While voting tendencies do change at the national level due to economic conditions or a change in population, this certainly one of the more unique aspects of state legislative elections.
  
  As is the case with senatorial elections, a candidate’s chances of winning, depending on the affiliated party, can be heavily affected by the performance of other offices held by a politician in the same (or different) party.  For example, one variable is the midterm penalty, which can be seen in the us_results dataset below.  This variable is coded by a -1, 0, or 1 (with respect to a Democratic candidate) and implies that the candidate in the party opposite of the President (Republican or Democrat) is highly favored to win if that election takes place halfway (two years) through a Presidential term.  Additionally, there are other variables in the us_results dataset with potentially important information such as approval rating.  In short, the outcome of state legislative elections can be influenced by the performance of candidates in national (and state) offices and the subsequent preferences in following elections.

  Our forecasting goal involves building a model that uses different covariates to predict the probability that a given California district will elect a democratic official in a two-party system. These covariates involve data from two different levels: the national and legislative district levels. Potential covariates involving data from the national level include presidential approval, strength of the economy and others. Some potential covariates involving data from the legislative district level include incumbency status of the seat up for election, vote totals from each party, and others. The achievement of such a model involves obtaining (messy) data from various sources. Although you will not be asked to build and test a model (certainly forecasting elections involve higher level modeling), the following exercises will help you understand the process of model building, starting from importing and cleaning messy datasets.

* How much does incumbency affect election results?
    + Do incumbents generally recapture their seat?
* Do districts usually vote according to the same party for which they voted previously?
* Is voting behavior for national offices (such as offices of the president, governor, and US senators) a good predictor for voting behavior at the state legislative level?

The exercises below will help answer these questions. To complete the exercises, you will need to use the RCurl, dplyr, tidyr, and ggplot2.

## Exercises

### Exericse 1

Import the data using the following code

```{r, echo = TRUE}
econ <-
  read.csv(text = getURL("https://raw.githubusercontent.com/ds-elections/state-forecasting-alpha/master/CleanDeliverableData/econ_state.csv"),
           header = TRUE)[,-1]
leg_results <-
  read.csv(text = getURL("https://raw.githubusercontent.com/ds-elections/state-forecasting-alpha/master/CleanDeliverableData/leg_results.csv"), header = TRUE)[,-1]
us_results <-
  read.csv(text = getURL("https://raw.githubusercontent.com/ds-elections/state-forecasting-alpha/master/CleanDeliverableData/us_results.csv"), header = TRUE)[,-1]
```

The leg_results and us_results datasets are fully clean. Each observation shows voting totals and covariate data for a single chamber district. The econ dataset shows the national average real personal income after each quarter since 1965.

* Merge leg_results and us_results together.
* Tidy the econ dataset so that each observation shows quarterly changes in a year. Then calculate for a weighted change in income so that the highest weight is attributed to the most recent quarterly change, and the lowest to the least recent quarterly change.
    + (Hint: Elections are in November, so the most recent quarterly change occured from March through June)
* Merge the econ dataset with the merged results dataset to reproduce the dataset below.

```{r, echo = FALSE}
results <- full_join(leg_results, us_results, by = c("year", "chamber", "District"))

econ <- econ %>%
  spread(key = Quarter, value = Income) %>%
  mutate(I_IIpercent = (II-I)/I,
         II_IIIpercent = (lag(III)-lag(II))/lag(II),
         III_IVpercent = (lag(IV)-lag(III))/lag(III),
         IV_Ipercent = (I - lag(IV))/lag(IV)) %>%
  mutate(perc_change = (3*I_IIpercent +
                        4*II_IIIpercent +
                        III_IVpercent +
                        2*IV_Ipercent)/10) %>%
  rename(year = Year)

results <- left_join(results, econ, by = "year")

results <- results %>%
  mutate(incumb_party_name = ifelse(incumbentparty == 1,
                                    "Democrat",
                                    ifelse(incumbentparty == 0,
                                           "No incumbent",
                                           "Republican"))) %>%
  arrange(year, chamber, District)
```

```{r}
head(results)
```


### Exercise 2

* Reproduce the plot below to capture how incumbency is related to the outcome of legislative elections. To understand the dataset, we need to define a few variables:
    + Dem = Total votes awarded to the democractic legislative candidate
    + Repub = Total votes awarded to the republican legislative candidate
    + incumbentparty = -1 if the election includes a republican incumbent, 1 if the                          election includes a democratic incumbent, and 0 if no incumbent                        is included

```{r, echo = FALSE}
plot1 <- ggplot(results, aes(y = Dem, x = Repub)) +
  geom_point(mapping = aes(col = incumb_party_name)) +
  scale_colour_manual(values = c("blue", "grey", "red")) +
  geom_abline(intercept = 0, slope = 1) +
  ylab("Vote count for the Democratic Candidate") +
  xlab("Vote count for the Republican Candidate") +
  labs(col = "Incumbency status") +
  ggtitle("The incumbency effect")
```

```{r}
plot1
```

* Other than a change in political views, what do you think could be a cause for the incumbent party losing the election (i.e. a Republican receiving more votes in an election where the Democratic party was the incumbent party)? Does the data account for this? (hint: think about the property of legislative districts)

* The advantage of being an incumbent seems to diminish over time. How would you manipulate the plot to verify this trend?